---
title: "Figure 1: characterization of clusters"
output: html_notebook
---


```{r config, echo=FALSE, message=FALSE}
rm(list = ls())
suppressPackageStartupMessages({
  library(Seurat)
  library(tidyverse)
  library(openxlsx)
  library(stringr)
  library(RColorBrewer)
  library(gridExtra)
  library(cowplot)
  library(ggbeeswarm)
  library(ggsignif)
  library(ComplexHeatmap)
  library(circlize)
})
source("utils.R")

```

load data
```{r}
so_T <- readRDS("../../data/scRNAseq_10x/seuratObject_T_cells.rds")
so_nT <- readRDS("../../data/scRNAseq_10x/seuratObject_non_T_cells.rds")
```



## A. cartoon of study (Ania)


## B. data overview: time of sac vs. tumor size

```{r}
plt_df <- so_T@meta.data %>% 
  group_by(time_point, tumor_size, sample_num_new) %>% 
  tally(name = "n_cells")
lfit <- lm(tumor_size ~ time_point, data = plt_df)
lsum <- summary(lfit)
lsum$r.squared

cor(plt_df$time_point, plt_df$tumor_size, method = "spearman")
# [1] 0.8562886

p <- ggplot(plt_df, aes(x = time_point, y = tumor_size)) +
  # geom_smooth(method = "lm", formula = y ~ x, size = 0.5, linetype = 2, se = T) +
  geom_smooth(method = 'loess', formula = 'y ~ x', size = 0.5, linetype = 2, se = F) +
  geom_jitter(shape = 21, fill = "black", width = 0.1) +
  annotate("label", x = 10, y = 280, label = expression(paste("Spearman ", rho, "=0.86")), size = 4) +
  labs(x = "Time (days)", y = expression(paste("Tumor size (", mm^{2}, ")"))) +
  scale_x_continuous(breaks = c(9, 10, 11, 12, 13, 14, 15, 16)) +
  # scale_y_log10() +
  theme_classic() +
  theme(axis.title = element_text(size = 14, colour = "black"),
        axis.text = element_text(size = 12, colour = "black"))
set.seed(121)
pdf(paste0("B_scatter_size_vs_time_", Sys.Date(), ".pdf"), width = 5, height = 3)
p
dev.off()
```



## C. T cell clusters
UMAP of T cell clusters.

```{r}
## UMAP of clusters
center_df <- so_T@meta.data %>% 
  group_by(cluster_annotation, louvain) %>%
  summarise(center_x = mean(UMAP_1), 
            center_y = mean(UMAP_2)) %>% 
  mutate(label = louvain)

order_cluster_T <- sub("\\..*", "", names(color_cluster_annotation_T)) %>% sub("^.*T", "", .) %>% as.integer() %>% order()

p1 <- so_T@meta.data %>% 
  mutate(cluster_annotation = factor(cluster_annotation, levels = names(color_cluster_annotation_T)[order_cluster_T])) %>% 
  ggplot() +
  geom_point(aes(x = UMAP_1, y = UMAP_2, color = cluster_annotation), size = 0.01) +
  geom_label(data = center_df, aes(x = center_x, y = center_y, label = label), 
             size = 3, vjust = .5, fontface = "bold", lineheight = 0.8, fill = "white", alpha = 0.8, label.size = NA) +
  scale_color_manual(values = color_cluster_annotation_T) +
  labs(color = "", title = "T Cell Clusters") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 8),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.height = unit(0.1, "in"),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA))
legend <- get_plot_component(p1, 'guide-box-right', return_all = TRUE)

pdf(paste0("C_UMAP_T_clusters_", Sys.Date(), ".pdf"), width = 4, height = 3)
p1 + theme(legend.position = "none")
ggdraw(legend)
dev.off()
```


## D. Dotplot to show the expression (z-score) of marker genes that Davide selected to annotate the T cell clusters.

```{r}
## read and process cluster markers
m_T <- readWorkbook(wb, sheet = "T_markers_parsed")
setdiff(m_T$gene, rownames(so_T))
markers_T <- m_T %>% 
  filter(!duplicated(gene)) %>% 
  filter(gene %in% rownames(so_T)) %>% 
  mutate(cluster = factor(cluster, levels = unique(m_T$cluster)))
# check if all markers are present in the dataset

library(Matrix)
plt_df <- data.frame(
  cluster = so_T$cluster_name,
  t(so_T[["RNA"]]$data[markers_T$gene,])
) %>% gather(key = "gene", value = "expr", -cluster) %>% 
  group_by(cluster, gene) %>% 
  summarise(avg_expr = mean(expr),
         pct_expr = 100 * mean(expr > 0)) %>% 
  group_by(gene) %>% 
  mutate(avg_expr_z = scale(avg_expr)) %>% 
  ungroup() %>% 
  mutate(gene = factor(gene, levels = (markers_T$gene)),
         cluster = factor(cluster, levels = levels(so_T$cluster_name) %>% rev),
         gene_cluster = markers_T$cluster[match(gene, markers_T$gene)] %>% factor(levels = rev(levels(cluster))))

p_T <- ggplot(plt_df, aes(x = gene, y = cluster, size = pct_expr, fill = avg_expr_z)) +
  geom_point(shape = 21) +
  scale_fill_gradientn(colors = brewer.pal(11, "RdBu") %>% rev) +
  scale_size(range = c(0, 6)) +
  scale_y_discrete(labels = levels(so_T$cluster_name) %>% rev) +
  theme_classic() +
  labs(x = "", y = "", size = "Percent expressed", fill = "Z-normalized\naverage expression") +
  facet_grid(~gene_cluster, scales = "free", space = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black", face = "italic"),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text = element_blank())
  
p_T



## save to pdf
pdf(file = paste0("D_dotplot_T_cluster_markers_", Sys.Date(), ".pdf"), width = 9, height = 5)
p_T
dev.off()
```





## E. Non-T cell clusters
UMAP of non-T cell clusters
```{r}
## UMAP of clusters
center_df <- so_nT@meta.data %>% 
  group_by(annotation, louvain) %>%
  summarise(center_x = mean(UMAP_1), 
            center_y = mean(UMAP_2)) %>% 
  mutate(label = louvain)
center_df$center_x[center_df$louvain == 12] <- -1.3

order_cluster_nT <- sub("\\..*", "", names(color_cluster_annotation_nT)) %>% sub("^.*T", "", .) %>% as.integer() %>% order()
p1 <- so_nT@meta.data %>% 
  mutate(cluster_annotation = factor(cluster_annotation, levels = names(color_cluster_annotation_nT)[order_cluster_nT])) %>% 
  ggplot() +
  geom_point(aes(x = UMAP_1, y = UMAP_2, color = cluster_annotation), size = 0.01) +
  geom_label(data = center_df, aes(x = center_x, y = center_y, label = label), 
             size = 3, vjust = .5, fontface = "bold", lineheight = 0.8, fill = "white", alpha = 0.8, label.size = NA) +
  scale_color_manual(values = color_cluster_annotation_nT) +
  labs(color = "", title = "Non-T Cell Clusters") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 8),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.height = unit(0.1, "in"),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA))
legend <- get_plot_component(p1, 'guide-box-right', return_all = TRUE)

pdf(paste0( "E_UMAP_Non-T_clusters_", Sys.Date(), ".pdf"), width = 4, height = 3)
p1 + theme(legend.position = "none")
ggdraw(legend)
dev.off()

```


<!-- Heatmaps to show the expression (z-score) of marker genes that Davide selected to annotate the non-T cell clusters. -->

<!-- ```{r} -->
<!-- ## read and process cluster markers -->
<!-- m_nT <- readWorkbook(wb, sheet = "nonT_markers") -->
<!-- markers_nT <- gather(m_nT, key = "cluster", value = "gene") %>%  -->
<!--   filter(gene != "" & !is.na(gene)) %>% -->
<!--   mutate(cluster = factor(cluster, levels = ann_nT$cluster)) %>%  -->
<!--   arrange(cluster) %>%  -->
<!--   filter(!duplicated(gene)) -->
<!-- # parse gene names to match the seurat object -->
<!-- markers_nT$gene <- rownames(so_nT)[match(toupper(markers_nT$gene), toupper(rownames(so_nT)))] -->
<!-- # check if all markers are present in the dataset -->
<!-- stopifnot(!any(is.na(markers_nT$gene))) -->

<!-- color_cluster_a_nT <- color_cluster_nT %>% `names<-`(., value = ann_nT$annotation) -->
<!-- p_nT <- DoHeatmap(so_nT, features = markers_nT$gene,  -->
<!--                  group.by = "annotation",  -->
<!--                  group.colors = color_cluster_a_nT, -->
<!--                  size = 3, raster = TRUE, angle = 60) + -->
<!--   scale_fill_gradient2(low = "magenta", mid = "black", high = "yellow", breaks = seq(-2.5, 2.5, length.out = 5)) + -->
<!--   guides(color = FALSE) + -->
<!--   theme(axis.text.y = element_text(size = 9), -->
<!--         plot.margin = unit(c(1.5,0,0,0), "cm")) -->

<!-- ## save to pdf -->
<!-- pdf(file = paste0(dir_out, "D_heatmap_Non-T_cluster_markers_", today, ".pdf"), width = 8, height = 9) -->
<!-- p_nT -->
<!-- dev.off() -->
<!-- ``` -->


## F. Dotplot to show the expression (z-score) of marker genes that Davide selected to annotate the T cell clusters.

```{r}
## read and process cluster markers
m_nT <- readWorkbook(wb, sheet = "nonT_markers_parsed")
setdiff(m_nT$gene, rownames(so_nT)) %>% cat(sep = '\n')
markers_nT <- m_nT %>% 
  filter(!duplicated(gene)) %>% 
  filter(gene %in% rownames(so_nT)) %>% 
  mutate(cluster = factor(cluster, levels = unique(m_nT$cluster)))


library(Matrix)
plt_df <- data.frame(
  cluster = so_nT$cluster_name,
  t(so_nT[["RNA"]]$data[markers_nT$gene,]),
  check.names = FALSE
) %>% 
  filter(cluster != "Non-T12") %>% 
  gather(key = "gene", value = "expr", -cluster) %>% 
  group_by(cluster, gene) %>% 
  summarise(avg_expr = mean(expr),
         pct_expr = 100 * mean(expr > 0)) %>% 
  group_by(gene) %>% 
  mutate(avg_expr_z = scale(avg_expr)) %>% 
  ungroup() %>% 
  mutate(gene = factor(gene, levels = (markers_nT$gene)),
         cluster = factor(cluster, levels = levels(so_nT$cluster_name) %>% rev),
         gene_cluster = markers_nT$cluster[match(gene, markers_nT$gene)] %>% factor(levels = rev(levels(cluster))))

p_nT <- ggplot(plt_df, aes(x = gene, y = cluster, size = pct_expr, fill = avg_expr_z)) +
  geom_point(shape = 21) +
  scale_fill_gradientn(colors = brewer.pal(11, "RdBu") %>% rev) +
  scale_size(range = c(0, 6)) +
  # scale_y_discrete(labels = levels(so_nT$cluster_name) %>% rev) +
  theme_classic() +
  labs(x = "", y = "", size = "Percent expressed", fill = "Z-normalized\naverage expression") +
  facet_grid(~gene_cluster, scales = "free", space = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black", face = "italic"),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text = element_blank())
  
p_nT



## save to pdf
pdf(file = paste0("F_dotplot_nT_cluster_markers_", Sys.Date(), ".pdf"), width = 12, height = 5)
p_nT
dev.off()
```


## G. CD8 cluster trajectory

```{r}
cd8 <- readRDS("../../timecourse_sc_RNAseq_10x/2_CD8_focused/cd8.rds")
cd8_cluster_vec <- c("T3", "T0", "T1", "T7", "T4", "T9")
color_cluster_annotation_cd8 <- color_cluster_annotation_T[match(cd8_cluster_vec, names(color_cluster_T))]
cd8$cluster_annotation <- factor(cd8$cluster_annotation, levels = names(color_cluster_annotation_cd8))


lineage <- readRDS("../../timecourse_sc_RNAseq_10x/2_CD8_focused/slingshot_lineages.rds")
# sce <- readRDS("../../timecourse_sc_RNAseq_10x/2_CD8_focused/cd8_slingshot_sce.rds")
# tmp <- SlingshotDataSet(sce)
# tmp@adjacency

center_df <- data.frame(cd8@meta.data %>% select(annotation), cd8@reductions$umap@cell.embeddings) %>% 
  group_by(annotation) %>%
  summarise(center_x = mean(umap_1), 
            center_y = mean(umap_2))
lineage_df_ls <- lapply(lineage, function(x) {
  center_df %>% filter(annotation %in% x) %>% mutate(annotation = factor(annotation, levels = x)) %>% arrange(annotation)
})



p1 <- DimPlot(cd8, group.by = "cluster_annotation", reduction = "umap", label = FALSE) +
  geom_point(data = center_df, mapping = aes(x = center_x, y = center_y), size = 3) +
  geom_path(data = lineage_df_ls$Lineage1, mapping = aes(x = center_x, y = center_y), linewidth = 0.5) +
  geom_path(data = lineage_df_ls$Lineage2, mapping = aes(x = center_x, y = center_y), linewidth = 0.5) +
  scale_color_manual(values = color_cluster_annotation_cd8) +
  labs(color = "", title = "T Cell Clusters", x = "UMAP_1", y = "UMAP_2") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 8),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.height = unit(0.1, "in"),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA))
legend <- get_plot_component(p1, 'guide-box-right', return_all = TRUE)

pdf(paste0( "G_CD8_trajectory_", Sys.Date(), ".pdf"), width = 4, height = 3)
p1 + theme(legend.position = "none")
ggdraw(legend)
dev.off()

```



## H. UMAP and violin plots of CD8 signatures.
```{r}
sigs_to_plot <- c("Progenitor vs. terminally exhausted (Tumor, bulk)" = "Stem-like signature", 
                  "Short-lived effectors signature" = "Short-lived effectors signature",
                  "Terminally exhausted vs. Progenitor (Tumor, bulk)" = "Terminal dysfunction signature", 
                  "Non_responder" = "Non-responder to ICB signature")
plist_umap <- list()
plist_vln <- list()
for (x in names(sigs_to_plot)) {
  plt_df <- data.frame(UMAP_1 = cd8@reductions$umap@cell.embeddings[,1],
                       UMAP_2 = cd8@reductions$umap@cell.embeddings[,2],
                       sigscore = cd8@meta.data[,x],
                       cluster_name = cd8$cluster_name %>% factor(levels = cd8_cluster_vec), 
                       annotation = cd8$annotation) %>% 
    mutate(sigscore_cap = pmax(sigscore, quantile(sigscore, 0.05)) %>% pmin(., quantile(sigscore, 0.95))) %>% 
    arrange(sigscore_cap)
  lmt <- ceiling(10*max(abs(range(plt_df$sigscore_cap))))/10
  plist_umap[[x]] <- ggplot(plt_df, aes(x = UMAP_1, y = UMAP_2, color = sigscore_cap)) +
    geom_point(size = 0.5) +
    coord_fixed() +
    # scale_color_gradientn(colors = c("#22578B", "#73A9D2", "white", "#FDB393", "#b22222"), limits = c(-lmt, lmt)) +
    scale_color_gradientn(colors = c("#22578B", "#73A9D2", "white", "#FDB393", "#b22222")) +
    labs(color = "Signature score") +
    theme_classic() +
    theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(color = "black", fill = NA))
  
  if (x == "Progenitor vs. terminally exhausted (Tumor, bulk)") {
    comparisons_ls <- list(c("T7", "T1"), c("T7", "T0"), c("T7", "T3"))
    step_increase <- 0.08
  } else if (x == "Short-lived effectors signature") {
    comparisons_ls <- list(c("T0", "T1"), c("T7", "T1"), c("T4", "T7"), c("T3", "T1"), c("T4", "T1"))
    step_increase <- c(0, 0.08, 0.08, 0.08, 0.08)
  } else if (x == "Terminally exhausted vs. Progenitor (Tumor, bulk)") {
    comparisons_ls <- list(c("T7", "T4"), c("T1", "T4"))
    step_increase <- 0.08
  } else if (x == "Non_responder") {
    comparisons_ls <- list(c("T4", "T9"), c("T4", "T7"), c("T1", "T4"))
    step_increase <- 0.08
  } else {
    stop("ERROR: Invalid signature name!")
  }
  
  
  plist_vln[[x]] <- ggplot(plt_df, aes(x = cluster_name, y = sigscore, color = cluster_name)) +
    geom_quasirandom(size = 0.5) +
    geom_boxplot(outliers = FALSE, fill = NA, color = "black", width = 0.1) +
    # coord_fixed(ratio = 6) +
    geom_signif(comparisons = comparisons_ls,
                step_increase = step_increase, color = "black", textsize = 2.8) +
    scale_color_manual(values = color_cluster_T) +
    labs(x = "", y = "Signature score") +
    theme_classic() +
    theme(legend.position = "none", plot.margin = margin(t = 0.1, r = 0.1, l = 0.1, b = 0.01, unit = 'in'),
          # plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
          axis.text.x = element_text(angle = 0, hjust = 0.5, color = "black", size = 12),
          axis.text.y = element_text(color = "black", size = 12),
          axis.title = element_text(color = "black", size = 14))
}

pdf(paste0("H_UMAP_violin_CD8_signatures_", Sys.Date(), ".pdf"), width = 10, height = 4)
for (i in names(plist_umap)) {
  grid.arrange(plist_umap[[i]], plist_vln[[i]], nrow = 1, widths = c(1, 0.4), top = textGrob(sigs_to_plot[i], gp = gpar(fontsize = 14)))
}
dev.off()
```



## I. TCR analysis

See TCR.Rmd


```{r}
meta_plate <- readRDS("../../timecourse_sc_RNAseq_smartseq2/meta_cd4_cd8_clonotype_2021-08-27.rds")
meta_sub <- meta_plate %>% 
  mutate(annotation = factor(names(color_annotation_T)[match(paste0("T", meta_plate$predicted.id), names(color_cluster_T))], levels = names(color_annotation_T))) %>% 
  filter(!is.na(clonotype_id), clone_size >= 3) %>% 
  arrange(-clone_size)
tab <- table(meta_sub$clonotype_id, meta_sub$annotation) 
pct <- (tab/rowSums(tab)*100) %>% as.data.frame() %>% spread(key = Var2, value = Freq) %>% column_to_rownames(var = "Var1")

idx_row <- do.call(order, -cbind(pct, rowSums(pct>0)))
plt_df_total <- data.frame(Total = rowSums(tab)[idx_row])
plt_mtx_pct <- pct[idx_row,]

df_annot_col <- data.frame(
  cluster = names(color_annotation_T)
) %>% `rownames<-`(., value = colnames(pct))
# color_annotation_T_ss2 <- color_annotation_T
# names(color_annotation_T_ss2) <- df_annot_col$cluster

annot_col <- columnAnnotation(cluster = df_annot_col$cluster, 
                              col = list(cluster = color_annotation_T),
                              show_legend = c("col" = FALSE), 
                              gp = gpar(fontsize = 9))

annot_row <- rowAnnotation(`Number\nof cells` = anno_barplot(plt_df_total$Total, axis_param = list(direction = "reverse"), border = TRUE),
                           width = unit(0.8, "in"), gp = gpar(fontsize = 9))

hm <-Heatmap(plt_mtx_pct,
  name = "Percent\nof cells",
  col = colorRamp2(breaks = seq(0, 100, length.out = 100), colors = colorRampPalette(c("white", "red"))(100)),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_names_side = "top",
  column_names_rot = 45,
  column_names_gp = gpar(fontsize = 9), 
  border = TRUE, 
  rect_gp = gpar(col = "grey", lwd = 0.01),
  top_annotation = annot_col,
  left_annotation = annot_row)


pdf(paste0("I_heatmap_clonotype_share_by_clusters_", Sys.Date(), ".pdf"), width = 4, height = 6)
draw(hm)
dev.off()


```


## S1A. UMAP color by time points and tumor sizes for T cells (related to Fig 1C)
```{r}
idx_shuffle_t <- sample(1:ncol(so_T), ncol(so_T), replace = FALSE)
## UMAP of clusters
p_t_time <- so_T@meta.data[idx_shuffle_t,] %>% 
  ggplot() +
  geom_point(aes(x = UMAP_1, y = UMAP_2, color = time_point), size = 0.01) +
  # scale_color_viridis_c(option = "D") +
  scale_color_gradientn(colors = brewer.pal(9, "BuPu")) +
  labs(color = "Time (days)") +
  coord_fixed() +
  # guides(color = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 8),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.height = unit(0.1, "in"),
        panel.background = element_blank())
# legend1 <- get_plot_component(p1, 'guide-box-right', return_all = TRUE)

p_t_size <-
  so_T@meta.data[idx_shuffle_t,] %>% 
  ggplot() +
  geom_point(aes(x = UMAP_1, y = UMAP_2, color = tumor_size), size = 0.01) +
  scale_color_gradientn(colors = brewer.pal(9, "OrRd")) +
  # scale_color_viridis_c(option = "B") +
  labs(color = "Tumor size (mm2)") +
  coord_fixed() +
  # guides(color = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 8),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.height = unit(0.1, "in"),
        panel.background = element_blank())

pdf(paste0("S1A_UMAP_by_time_size_T_", Sys.Date(), ".pdf"), width = 5, height = 3)
p_t_time
p_t_size
dev.off()
```


## S1B. UMAP color by time points and tumor sizes for non-T cells (related to Fig 1E)
```{r}
idx_shuffle_nt <- sample(1:ncol(so_nT), ncol(so_nT), replace = FALSE)
p_nt_time <- so_nT@meta.data[idx_shuffle_nt,] %>% 
  ggplot() +
  geom_point(aes(x = UMAP_1, y = UMAP_2, color = time_point), size = 0.01) +
  # scale_color_viridis_c(option = "D") +
  scale_color_gradientn(colors = brewer.pal(9, "BuPu")) +
  labs(color = "Time (days)") +
  coord_fixed(ratio = 0.8) +
  # guides(color = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 8),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.height = unit(0.1, "in"),
        panel.background = element_blank())
# legend1 <- get_plot_component(p1, 'guide-box-right', return_all = TRUE)

p_nt_size <-
  so_nT@meta.data[idx_shuffle_nt,] %>% 
  ggplot() +
  geom_point(aes(x = UMAP_1, y = UMAP_2, color = tumor_size), size = 0.01) +
  scale_color_gradientn(colors = brewer.pal(9, "OrRd")) +
  # scale_color_viridis_c(option = "B") +
  labs(color = "Tumor size (mm2)") +
  coord_fixed(ratio = 0.8) +
  # guides(color = guide_legend(override.aes = list(size = 3))) +
  theme(plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        axis.title = element_text(size = 8),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.text = element_text(size = 8),
        legend.key.height = unit(0.1, "in"),
        panel.background = element_blank())

pdf(paste0("S1B_UMAP_by_time_size_nonT_", Sys.Date(), ".pdf"), width = 5, height = 3)
p_nt_time
p_nt_size
dev.off()
```

## S1C. UMAP plots of genes and signatures.
```{r}
source("~/project/utils/map_genes/map_human_mouse_genes.R")
cell_cycle_genes <- union(human_to_mouse(cc.genes.updated.2019$s.genes), human_to_mouse(cc.genes.updated.2019$g2m.genes))
so_T <- AddModuleScore(so_T, features = list(cell_cycle_genes), name = "cell_cycle")
colnames(so_T@meta.data) <- sub("cell_cycle1", "Cell cycle", colnames(so_T@meta.data))

features_to_plot <- list(
  "T" = c("Cd3e", "Klra7", "Cd8a", "Cd4", "Pdcd1", "Havcr2", "Foxp3", "Cell cycle"),
  "nT" = c("Cd14", "Cd24a", "Cd209a", "Ccr7", "Cd19", "Ncr1", "Xcr1", "Bst2")
)

plist <- lapply(names(features_to_plot), function(x) {
  if (x == "T") {so <- so_T} else {so <- so_nT}
  lapply(features_to_plot[[x]], function(f) {
    plt_df <- so@meta.data %>% select(UMAP_1, UMAP_2)
    if (f %in% rownames(so)) {
      plt_df$value <- as.numeric(so[["RNA"]]@data[f,])
      lab_value <- "Expression"
    } else {
      plt_df$value <- so@meta.data[,f]
      lab_value <- "Signature\nscore"
    }
    plt_df %>% 
      arrange(value) %>% 
      ggplot() +
      geom_point(aes(x = UMAP_1, y = UMAP_2, color = value), size = 0.01) +
      scale_color_gradientn(colors = brewer.pal(9, "Blues")[-1]) +
      labs(color = lab_value, title = f) +
      theme(plot.title = element_text(size = 10, hjust = 0.5, face = "italic"),
          axis.title = element_text(size = 8),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 8),
          legend.key.width = unit(0.1, "in"),
          legend.key.height = unit(0.1, "in"),
          panel.background = element_blank(),
          panel.border = element_rect(color = "black", fill = NA))
  })
})

pdf(paste0("S1C_UMAP_features_", Sys.Date(), ".pdf"), width = 6, height = 7)
grid.arrange(grobs = plist[[1]], ncol = 2)
grid.arrange(grobs = plist[[2]], ncol = 2)
dev.off()
```



## S1D. violin plot compare cell cycle signature scores in T4, T7, T2 and T8.
Plot selected pathways.
```{r}
clusters_to_plot <- c("T4", "T7", "T2", "T8")
plt_df <- so_T@meta.data %>% 
  filter(cluster_name %in% clusters_to_plot)
p_cc <-
  ggplot(plt_df, aes(x = cluster_annotation, y = `Cell cycle`)) +
  geom_quasirandom(aes(color = cluster_annotation)) +
  geom_boxplot(outliers = FALSE, fill = NA, color = "black", width = 0.1, size = 1) +
  geom_signif(comparisons = list(c("T7. CD8+ PD1+ TIM3+ 2", "T4. CD8+ PD1+ TIM3+ 1"),
                                 c("T2. CD4+ Treg 1", "T8. CD4+ Treg 2")), margin_top = 0.1) +
  scale_color_manual(values = color_cluster_annotation_T) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  theme_classic() +
  labs(x = "", y = "Signature score", color = "", title = "Cell cycle") +
  theme(axis.text.x = element_text(color = "black", angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "none")

pdf(paste0("SD_violin_cell_cycle_score_", Sys.Date(), ".pdf"), width = 3.5, height = 4)
p_cc
dev.off()
```



## S1E. T5 vs. T6
Plot selected pathways.
```{r}
plt_df <- read.xlsx("GSEA_T5_vs_T6_selected.xlsx", sheet = "to_plot") %>% 
  filter(!grepl("^Non-Tfh", Name_in_plot)) %>% 
  mutate(enriched_in = factor(enriched_in, levels = c("T_5", "T_6"), labels = c("T5. CD4+ Tconv 1", "T6. CD4+ Tconv 2")))
plt_df$Name_in_plot <- factor(plt_df$Name_in_plot, levels = plt_df$Name_in_plot, labels = sub(" \\(", "\n\\(", plt_df$Name_in_plot))
plt_df$pvalue_label <- paste0(" p=", formatC(plt_df$pvalue, digits = 2, format = "e"), " ")
p <- ggplot(plt_df, aes(x = Name_in_plot, y = NES)) +
  geom_bar(aes(fill = enriched_in), stat = "identity", color = "black", width = 0.7) +
  geom_text(aes(label = pvalue_label, hjust = NES > 0), y = 0, size = 3) +
  geom_hline(yintercept = 0, color = "black") +
  geom_vline(xintercept = 0, color = "black") +
  coord_flip() +
  scale_fill_manual(values = color_cluster_annotation_T) +
  theme_minimal() +
  labs(y = "Normalized enrichment score", x = "", fill = "Enriched in") +
  theme(axis.text = element_text(color = "black"))
pdf(paste0("S1E_barplot_T5_vs_T6_enrichment_", Sys.Date(), ".pdf"), width = 7, height = 3.4)
p
dev.off()
```


## S1F. SS2 cell label prediction.
```{r}
meta_plate <- readRDS("../../timecourse_sc_RNAseq_smartseq2/meta_cd4_cd8_clonotype_2021-08-27.rds")
plt_df <- meta_plate %>% 
  mutate(annotation = factor(names(color_annotation_T)[match(paste0("T", meta_plate$predicted.id), names(color_cluster_T))], levels = names(color_annotation_T))) %>% 
  group_by(orig.ident, annotation) %>% 
  tally()

p <- ggplot(plt_df, aes(x = annotation, y = n, fill = orig.ident)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = c("CD4" = alpha("red", 0.5), "CD8" = alpha("black", 0.5))) +
  labs(x = "Predicted cell type", y = "Number of cells", fill = "FACS sort\ncell type") +
  theme_classic() +
  theme(axis.text.x = element_text(color = "black", angle = 45, hjust = 1))
pdf(paste0("S1F_barplot_predicted_celltype_", Sys.Date(), ".pdf"), width = 5, height = 4)
p
dev.off()
```





